---
title: "SVM Classification"
author: "Jasfer David Gabriel"
output: html_document
---

------------------------------------------------------------------------

#### Libraries

```{r warning=FALSE, message=FALSE}
#Loading necessary Libraries
library(e1071)
```

------------------------------------------------------------------------

#### Loading and Viewing Dataset

```{r}
data(iris)
```

-   The `data(iris)` loads the built-in iris dataset into your R environment.

```{r}
#Summary of dataset
str(iris)
```

-   The `str(iris)` provides the summary of the iris dataset.
-   The dataset has a total of 150 records and 5 variables.

```{r}
attach(iris)
```

-   The `attach(iris)` makes the columns of the `iris` dataset directly accessible by name, without using the `$` operator.

```{r}
#Accessing columns in iris dataset without using $ operator
Species[!duplicated(Species)]
```

-   The `iris` flowers dataset has three species setosa, versicolor, virginca.
-   `duplicated(species)` combined with `!` operator gives the unique values in `species` column.

------------------------------------------------------------------------

#### Classification Model : Support vector Machine(SVM)

Fit `svm` model to predict the species of an iris flower.

```{r}
# Fit SVM model
model <- svm(Species ~., data = iris) 
```

-   `svm()` function is used to train an SVM model.
-   `Species ~ .`: A formula where `Species` is the target variable, and the `.` means all other variables in the dataset are used as predictors.
-   `data = iris`: Specifies the iris dataset as the source of both the target and predictors.

```{r}
# Model summary
summary(model)
```

-   `SVM-Type: C-classification` means the model is used for classification, not regression.

-   `SVM-Kernel: radial` means the model Uses the Radial Basis Function (RBF) kernel, good for non-linear classification problems.

-   `Cost: 1` is the default value that controls the penalty for misclassification.

-   `Number of Support Vectors: 51`

    -   The model found 51 support vectors out of 150 observations.
    -   That means only these 51 points are actively defining the classification boundaries between the 3 species.
    -   The remaining 99 points are correctly classified but are not close enough to the boundary to be considered support vectors.

-   `Number of Classes: 3` means the model classifies iris species into three categories

    -   setosa(8)

    -   versicolor(22)

    -   virginica(21)

Model Testing

```{r}
# Model Testing
prediction <- fitted(model)
```

-   This extracts the fitted values (predicted classes) from the trained SVM `model` and stores them in the `prediction` variable.

> `fitted(model)` gives predictions on the training data, not on new or test data.

Calculating Accuracy of the Model

```{r}
y <- Species # target variable
table(prediction, y)
```

-   `table(prediction, y)` creates a confusion matrix comparing:`prediction`: the predicted class labels from your SVM model `(fitted(model))` and `y`: the actual class labels (Species)

-   Accuracy:

    -   Setosa: Perfect classification - all 50 correctly classified.
    -   Versicolor:48 correctly predicted.2 mistakenly predicted as virginica.
    -   Virginica:48 correctly predicted.2 mistakenly predicted as versicolor.

```{r}
#calculating Accuracy
sum(prediction==y)/length(y) * 100
```

-   The model performs very well, with an overall accuracy of 97.33%

Visualizing SVM model

```{r}
plot(cmdscale(dist(iris[,-5])),
     col = as.integer(iris[,5]),
     pch = c("o","+")[1:150 %in% model$index + 1],
     xlab = "",
     ylab = "",
     main = "SVM Classification with Support Vectors")
```

-   Each point represents a flower sample from the iris dataset.
-   Colors: Represent the 3 different iris species setosa(Black),versicolor(Red) and virginica(Green).
-   Shape:
    -   `"o"` = Regular data point
    -   `"+"` = Support vector
-   The SVM model learns the species boundaries very well, especially for setosa(Black).
-   The overlap between versicolor(Red) and virginica(Green) indicates they are more similar and harder to separate.

> This visual helps show which points the model relied on most (support vectors) and how well the classes are separated.

------------------------------------------------------------------------

##### Recommendations:

-   Use this model for botanical classification tasks, especially when species show clear visual or measurable differences in physical features.

-   For real-world deployment, test the model on new data (not just training set).

-   To improve the model further:

    -   Use cross-validation to avoid overfitting.

##### Variable Suggestions for Extended Use:

-   In a real flower classification system, consider adding:
    -   Geographic location (region/country of origin)
    -   Soil type, sunlight exposure or blooming season
    -   Environmental data (from external sources)

------------------------------------------------------------------------

#### Regression Model : Support vector Machine(SVM)

```{r}
x <- seq(0.1, 5, by = 0.05)
length(x)
```

-   Creates a sequence of 99 values from 0.1 to 5 with 0.05 increments.

```{r}
y <- log(x) + rnorm(x, sd = 0.2)
```

-   Computes the natural log of x, then adds random noise using rnorm() with a standard deviation of 0.2.

-   This simulates real-world noisy observations.

    -   Example output:
        -   log(1) = 0,might become 0.15 due to noise.
        -   log(2) \~ 0.69,could become 0.55 or 0.90 depending on randomness.

Fit the Model

```{r}
mod <- svm(x,y)
summary(mod)
```

-   Indicates a regression model was trained using vectors x and y.
-   Out of 99 training points, 67 are support vectors.These are the most influential points used to define the regression function.

```{r}
pred <- predict(mod,x)
```

-   Predicts the response(y) using the model for each value of x.

Visualize the model

```{r}
plot(x, y)
points(x, log(x), col = 2)
points(x, pred, col = 4)
```

-   This plot shows how the Support Vector Machine - Regression model performs compared to the actual logarithmic function and noisy data.
-   Black circles Noisy data points (y) = log(x) + random noise
-   Red circles True log(x) function (without noise)
-   Blue circles predicted values (pred) using svm()

Predictions (Blue):

-   The blue curve smoothly follows the true red curve.

-   It stays close to log(x) and filters out random noise.

-   It doesn’t overfit to every noisy point, which is ideal in regression.

True Function (Red):

-   Serves as the benchmark for evaluating model performance.

-   Both red and blue are close - showing the model learned the right pattern.

Noise Handling:

-   The black dots (noisy y) fluctuate around the red curve.

-   Model effectively ignores most of this noise.

------------------------------------------------------------------------

##### Recommendations:

Use SVM regression when:

-   Data is non-linear

-   Data contains noise

------------------------------------------------------------------------

#### Unsupervised Learning : Support vector Machine(SVM)

```{r}
X <- data.frame(a = rnorm(1000), b = rnorm(1000))
attach(X)
```

-   Generates 1,000 random points for a and b, both from a standard normal distribution.

Fit the Model

```{r}
m <- svm(X, gamma = 0.1)
summary(m)
```

-   Uses one-classification SVM.
-   X is used as input, and the model tries to find the boundary around the main data cluster.
-   Out of 1000 points, 501 points are used as support vectors - they define the boundary of the learned region.
-   Number of Classes: 1,the model learns from a single group of normal data.

Model Testing

```{r}
newdata <- data.frame(a = c(0, 4), b = c(0, 4))
predict(m, newdata)
```

-   Tests whether the points (0,0) and (4,4) are inliers or outliers based on the learned boundary.
-   predict() returns TRUE (inlier) or FALSE (outlier).
-   (0,0) is inlier and (4,4) is outlier.

Visualize the Model

```{r}
plot(X, col = 1:1000 %in% m$index + 1, xlim = c(-5,5), ylim=c(-5,5))
points(newdata, pch = "+", col = 2, cex = 5)
```

-   Black circles: Points considered support vectors or inliers by the SVM.

-   Red circles: Data points that are outside the SVM decision boundary - potential outliers or non-support vectors.

-   Large red "+" markers:Mark the test points provided ((0, 0) and (4, 4)).

    -   (0, 0) lies near the center,classified as inlier.
    -   (4, 4) is far from the data,likely classified as outlier.

-   The SVM learned the core shape of the data.

-   The SVM model is successfully filtering out distant or low-density points.

-   It correctly flags unusual points (like (4, 4)) as potential anomalies.

------------------------------------------------------------------------

##### Recommendations:

Use One-Class SVM for:

-   Outlier Detection

-   Novelty Detection

------------------------------------------------------------------------
