---
title: "Text Mining on Dr.Martin Luther King's 'I Have a Dream' Speech"
author: "Jasfer David Gabriel"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

------------------------------------------------------------------------

## Introduction

-   This report performs text mining on Dr. Martin Luther King’s "I Have a Dream" speech.
-   The goal is to:
    -   clean the text
    -   Analyze word frequencies
    -   Generate visualizations such as a word cloud and bar plots to provide meaningful interpretations and recommendations based on the results.

------------------------------------------------------------------------

#### 1. Loading Libraries

```{r warning=FALSE, message=FALSE}
#Loading necessary Libraries
library(tm)
library(SnowballC)
library(wordcloud)
library(RColorBrewer)
```

-   These libraries provide tools for text processing, stemming and visualization.

------------------------------------------------------------------------

#### 2. Load and Inspect Data

```{r}
filePath <- "http://www.sthda.com/sthda/RDoc/example-files/martin-luther-king-i-have-a-dream-speech.txt"
text <- readLines(filePath) # Loads the text line by line
docs <- Corpus(VectorSource(text))  # converts text into a corpus
inspect(docs)


```

------------------------------------------------------------------------

#### 3. Text Cleaning

```{r warning=FALSE}
toSpace <- content_transformer(function(x, pattern) gsub(pattern, " ", x))
```

-   Defines a custom function `toSpace` that replaces any specified pattern with a space `" "`.

```{r warning=FALSE}
docs <- tm_map(docs, toSpace, "/")
docs <- tm_map(docs, toSpace, "@")
docs <- tm_map(docs, toSpace, "\\|")
```

-   Using `toSpace` replace pattern `(/, @, |)` with space.

```{R warning=FALSE}
docs <- tm_map(docs, content_transformer(tolower)) # converts all text to lower case
docs <- tm_map(docs, removeNumbers) # removes all numbers
docs <- tm_map(docs, removeWords, stopwords("english")) # remove common english words

# docs <- tm_map(docs, removeWords, c("blabla1", "blabla2")) # removes custom words

docs <- tm_map(docs, removePunctuation) # removes punctuations
docs <- tm_map(docs, stripWhitespace) # removes extra white spaces
docs <- tm_map(docs, stemDocument) # performs stemming
```

------------------------------------------------------------------------

#### 4.Create Document-Term Matrix

DocumentTermMatrix is a table that shows how often each word appears in each document.

-   We will create a matrix using the DocumentTermMatrix method by passing the corpus and then convert the object into matrix.

-   `sparsity` refers to the threshold of relative document frequency for a word, above which the word will be removed.

```{r}
dtm <- TermDocumentMatrix(docs)
dtm
```

------------------------------------------------------------------------

#### 5. Analysis of Word Frequencies

```{R}
m <- as.matrix(dtm)
v <- sort(rowSums(m), decreasing = TRUE)
d <- data.frame(word = names(v), freq = v, row.names = NULL)
```

The top 10 frequent words are,

```{R}
head(d,10)
```

```{r}
bp <- barplot(d[1:10, ]$freq,
              las = 2,
              names.arg = d[1:10, ]$word,
              col = "darkblue",
              main = "Top 10 Frequent Words",
              ylab = "Word Frequencies")

# Add text labels above the bars
text(x = bp,
     y = d[1:10, ]$freq,
     label = d[1:10, ]$freq,
     pos = 1,        # position of the lable in the bar
     cex = 0.8,      # font size
     col = "white")
```

-   The bar plot visually reinforces that these words are the rhetorical anchors of the speech.

Frequent Words

```{r}
#Returns all words that appear at least 7 times in your Document-Term Matrix (dtm).
findFreqTerms(dtm, lowfreq = 7)
```

-   words like “dream”, “freedom”, “day”, “let” and “ring” were confirmed to be highly frequent.

Frequent Co-occurence of Words

```{r}
#Returns words that frequently co-occur with the word "dream"
findAssocs(dtm, terms = "dream", corlimit = 0.5)
```

-   “dream” is strongly associated with:
    -   american, difficult, deeply, face, still, tomorrow, today (correlation ≥ 0.67–0.74).
    -   These associations show that the word “dream” is emotionally and thematically linked to struggle, hope, identity and the future.

------------------------------------------------------------------------

#### 6. Word Cloud Visualization

word cloud is a visual representation of text data where the size of each word indicates its frequency or importance in the source text.

```{r}
set.seed(1234)
wordcloud(words = d$word, freq = d$freq, min.freq = 1,
          max.words = 200, random.order = FALSE, rot.per = 0.35,
          colors = brewer.pal(8, "Dark2"))
```

The word cloud provides an immediate visual summary of the most prominent words.

Key observations:

-   “will”, “freedom”, “dream”, “day”, “ring” dominate visually, underscoring their rhetorical significance.

-    The stemming applied (e.g., “abl”, “togeth”,"everi") groups related words, though it may affect interpretability.

-   Words like “mountain”, “children” and “justice” also appear prominently, reflecting both metaphor and message.

------------------------------------------------------------------------

## Takeaway

-   The most frequent word is "will", highlighting its thematic importance throughout the speech. Indicating determination and future-oriented action.
-   Other prominent words such as "freedom", "dream", "together" reflect the core values of equality, unity and civil rights.

------------------------------------------------------------------------

## Recommendations

-   For Further Analysis:
    -   Add sentiment analysis to explore emotional tone across the speech.
    -   Use n-grams (4-grams/trigrams) to capture impactful phrases like “I have a dream” or “let freedom ring”.
    -   Apply the same text mining techniques to other famous speeches to compare themes, tone and linguistic patterns in leadership communication.

------------------------------------------------------------------------

## Conclusion

This analysis of Dr. Martin Luther King's "I Have a Dream" speech through text mining provides both visual and quantitative insights into its powerful messaging.
The repeated themes of will, dream, justice and unity emerge clearly through frequency analysis and visualization.

------------------------------------------------------------------------
